---
title: "multivariate"
author: "Haleh Mawson"
date: "2025-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to multivariate statistics ##

```{r packages}
library(Rmisc) #useful all-around package
library(tidyverse) #contains ggplot2, dplyr, etc.
library(vegan) #multivariate stats package
library(labdsv) #community ecology
```

## non-metric multidimensional scaling ##

If you have subsampled your true replicates, it is best to either sum or average the subsamples to avoid pseudoreplication. Random effects are not as easy in multivariate analyses as they are in GLMs, although you can restrict permutations in certain ways if need be.
```{r subplot summarizing}
data.sum <- data %>% 
  group_by(true_replicate_factors) %>% #groups you want to summarize
  summarize_at(vars(start_col:end_col), sum) %>% #columns that need to be summarized, often abundances
  ungroup() #removes grouping
```

The data then need to be converted into a distance matrix.
```{r community matrix}
dat.resp <- dat.sum %>% 
  select(start_col:end_col) %>% #just the species abundances
  vegtab(minval = 0.05 * nrow(dat.sum)) %>% # Remove rare species
  decostand("max") #relativize by max so that responses are on equal scales

dat.dist <- vegdist(dat.resp) #convert to Bray Curtis distance measure
```
 
```{r NMDS}
set.seed(42) #same result every run
dat.nmds <- metaMDS(comm = dat.dist, 
                    autotransform = FALSE,
                    k = 3, #three dimensional solution
                    maxit = 300, #maximum number of iterations
                    #to find local minimum
                    try = 40, #minimum number of tries 
                    trymax = 100) #maximum number of tries
                    #to find global minimum
print(dat.nmds)
```
The axes of an NMDS plot are unitless and purely relative. The interpretation is simple and straightforward as a result: two points that are closer to each other are more similar, two points far away are less similar.  

* Stress: 
  + < 0.05: Excellent representation, can be interpreted with confidence
  + < 0.10: Good representation
  + < 0.20: Good for general trends, but the details may be unreliable
  + > 0.20: not especially reliable
  
To graph and interpret the NMDS results, it's best to add back in your explanatory variables. 

```{r attach explanatory variables to NMDS points}
dat.points <- data.frame(dat.nmds$points) 
dat.exp <- dat.sum %>% select(exp_var1, exp_var2)
dat.inf <- cbind(dat.exp, dat.points)
#the two dataframes should still be the same length
```

```{r plot NMDS}
#plotting the first two axes of the ordination
ggplot(data = dat.inf,
       aes(x = MDS1, #NMDS axis 1
           y = MDS2, #NMDS axis 2
           color = exp_var1,
           shape = exp_var2)) + #first 2 most important explanatory vars
  geom_point(cex = 3) + #bump up the point size
  theme_bw() + #just a nice theme
  #then remove axis labels to prevent misinterpretation
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
  labs(caption = "Stress = 0.11, k = 3") #add in label with key info
  #helps show how confident we can be about any inferences from the graph
```

## PERMANOVA ##

## principle components analysis ##